\MakeUppercase{Linear Versus Centred Colouring via Pseudogrids}\thanks{This research was partly funded by NSERC.}

### 2025-09-19 lin-vs-cen.tex
Added keywords as requested by Wiley
Switched to PDF figures
Pffft

\title{\MakeUppercase{Linear versus centred chromatic numbers}\thanks{This research was partly funded by NSERC.}}
\author{Prosenjit~Bose\thanks{School of Computer Science, Carleton University},\quad Vida~Dujmović\thanks{Department of Computer Science and EE, University of Ottawa},\quad Hussein~Houdrouge\footnotemark[2],\quad Mehrnoosh~Javarsineh\footnotemark[2],\quad and Pat~Morin\thanks{School of Computer Science, Carleton University, \email{morin@scs.carleton.ca}}}

@@

@@

Any further improvement to \cref{kun_obrien_general2} will either require an improved Excluded Grid Theorem or an entirely new approach. However, no improvement to the Excluded Grid Theorem will sufficient to establish a linear relationship between the centred and the linear chromatic number.  Indeed, the best possible Excluded Grid Theorem would state that any graph of treewidth $k^2\log k$ contains an $\Omega(k)\times\Omega(k)$ grid minor \cite{robertson.seymour.ea:quickly}, and the preceding argument would only show that $\chicen(G)\in O((\chilin(G))^3\log(\chilin(G)))$.  Even the Excluded Grid Theorem for Planar Graphs states that any planar graph of treewidth $k$ contains an $\Omega(k)\times\Omega(k)$ grid minor
% \cite[(6.2)]{robertson.seymour.ea:quickly}.
\cite{robertson.seymour.ea:quickly}.
Combining this with the argument above and \cref{pseudogrid_lower_bound} shows only that, for any planar graph $G$, $\chicen(G)\in\Theta((\chilin(G))^2)$.

@@

\begin{figure}
  \begin{center}
    \includegraphics{figs/pseudogrid-2}
  \end{center}
  \caption{A $6\times 4$ pseudogrid.  Black vertices are contained in $P_v$ for some vertex $v$ of $G_{6\times 4}$ and white vertices are contained in $P_{vw}$ for some edge of $G_{6\times 4}$. Colour are used to distinguish between cases (Q1) (blue), (Q2) (pink), and (Q3) (gold).}
  \label{pseudogrid_fig}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{ccc}
      \includegraphics{figs/outline-2} &
      \includegraphics{figs/outline-3} &
      \includegraphics{figs/outline-4} \\
      $S_1$ & $S_1\cup S_2$ & $P$
    \end{tabular}
  \end{center}
  \caption{Constructing the sets $S_1$ and $S_2$ and the path $P$.}
  \label{outline}
\end{figure}
\begin{compactenum}

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{ccc}
      \includegraphics{figs/bad-example-1} &
      \includegraphics{figs/bad-example-2} &
      \includegraphics{figs/bad-example-3}
    \end{tabular}
  \end{center}
  \caption{Some examples of $3$- and $4$-element vertex subsets that cannot all be contained in a single path (that must also contain other vertices not pictured).}
  \label{bad_examples}
\end{figure}

@@

  \begin{figure}
    \begin{center}
      \begin{tabular}{ccc}
        \includegraphics[width=.3\textwidth]{figs/make_disjoint-1} &
        \includegraphics[width=.3\textwidth]{figs/make_disjoint-2} &
        \includegraphics[width=.3\textwidth]{figs/make_disjoint-3} \\
        $X_1$ & $X_2$ & $X_3$
      \end{tabular}
    \end{center}
    \caption{The proof of \cref{make_disjoint}. Disks are vertices in $S$ and crosses are the resulting elements of $X$. The large yellow boxes are $B_{r-1}(v)$ for some $v\in S$ and the small mauve and green boxes are $B_{p+1}(x)$ and $B_{p}(x)$, respectively, for some $x\in X$.}
    \label{make_disjoint_fig}
  \end{figure}
  Let $S_2$ be the subset of $S\setminus S_1$ containing those vertices that take part in a $2p$-pair.  For each $2p$-pair $v,w\in S_2$, there exists  $x\in V(\interior_r(G))$ such that $v,w\in \tilde{B}_{p}(x)$. We include $x$ in $X_2$ and say that $x$ \defin{covers} $v$ and $w$.

@@

\begin{proof}
  We begin by applying \cref{one_colour_per_object} to obtain the colouring $\phi:\VE(G_k)\to\varphi(V(G))\cup\{\perp\}$ in which $\phi^{-1}(\alpha)\ge d$ for each $\alpha\in\varphi(V(G))$.  Observe that it is now sufficient to find a $2|\varphi(V(G))|$-element subset $Q\subseteq\VE(\interior_r(G_k))$ such that
  \begin{compactenum}[(i)]
    \item $|\phi^{-1}(\alpha)\cap Q|= 2$ for each $\alpha\in\varphi(V(G))$ and
    \item $|\VE(G_r(\mu))\cap Q|\le 2$ for each $\mu\in Q$.
  \end{compactenum}
  Indeed, with such a $Q$ we can obtain $S$ by taking one vertex of colour $\phi(\mu)$ from $P_\mu$ for each $\mu\in Q$.

@@

  For each $\alpha\in X$, let $\Phi_\alpha:=\phi^{-1}(\alpha)\cap \left(\bigcup_{\mu\in Y}\VE(G_{2r+1}(\mu))\right)$.  In words, $\Phi_\alpha$ consists of all objects of colour $\alpha$ that were eliminated from consideration because they are too close to an element of $Y$. We claim that, for each $\alpha\in X$, $|\Phi_\alpha|\ge d-\vol{r}$.  Otherwise $|\phi^{-1}(\alpha)\setminus\Phi_\alpha| > \vol{r}$. Then, when the colour $\alpha$ was considered during the first round, we could have taken $\mu_1$ to be any object in $\phi^{-1}(\alpha)\setminus\Phi_\alpha$ and taken $\mu_2$ to be any object in $\phi^{-1}(\alpha)\setminus\Phi_\alpha\setminus B_{2r+1}(\mu_1)$. (Recall that $\vol{r}\ge |B_{2r+1}(\mu_1)|$, by definition.)  Thus, the first round would have been an $\alpha$-success, contradicting the fact that $\alpha\in X$.

@@

  Comparing \cref{weight_product,probability} we see that both quantities decrease exponentially in $|\Phi_\alpha|$ but \cref{weight_product} does so at a rate that can be controlled with $\tau$.
  Taking $\tau = {32\vol{7r}^2}$, and using \cref{weight_product,probability}, we find that
  \[
     \Pr\left(E_\alpha\right) \le w(E_\alpha)\prod_{\beta\in\Gamma_\alpha}(1-w(E_\beta))
  \]
  provided that
  \[   |\Phi_\alpha| \ge {32\vol{7r}}\big(\ln({32\vol{7r}}+1) + {2\vol{2r+1}}\ln(2|\Phi_\alpha|)\big) \enspace .
  \]
  The right hand side of this last equation is of the form $O(r^2\log r+r^4\log|\Phi_\alpha|)$.  Since $|\Phi_\alpha|\ge d-\vol{r}\in d- O(r^2)$, this can therefore be satisfied for some $d\in O(r^4\log r)$.

@@

We are also grateful to Omer Angel for pointing out that the problem on a $k\times k$ grid (as opposed to pseudogrid) has a simple solution, illustrated in \cref{omer}: Partition the grid into $2\times 2$ grids and repeatedly delete a subgrid if it contains a vertex whose colour is unique.  If this causes the grid to become disconnected, keep only the largest remaining component and continue.  This process deletes at most $4\epsilon k$ vertices.  For $\epsilon < 1/4$, this implies that the graph that remains at the end of thie process has at least $k^2/2$ vertices.  In particular, the graph that remains is non-empty and every colour that appears in this graph is used on at least two vertices.  Finally, this graph has a Hamiltonian path (which can be constructed by traversing the boundary of a certain tree contained in the dual graph).  Unfortunately, this solution breaks down immediately for pseudogrids since a pseudogrid is not necessarily Hamiltonian, even before removing some of its vertices and edges.

@@

  Although this is a simplification of the problem, we first describe how to solve it.  For this simpler problem, we may assume that each of $\mu_s$, $\mu_v$, $\mu_w$, and $\mu_t$ is an edge of $G_a$ since we can replace any vertex of $G_a$ with one of its incident edges. Refer to \cref{generic_path}. Beginning at $\mu_s$, $P_a$ can traverse the boundary of $G_a$ until reaching the first column $i$ that contains an endpoint of $\mu_v$ or $\mu_w$, then vertically in this column to collect $\mu_v$ (say). What happens next depends on whether or not $\mu_w$ also intersects column $i$.
  \begin{figure}[htbp]
    \begin{center}
      \begin{tabular}{cc}
        \includegraphics{figs/generic_path-1} &
        \includegraphics{figs/generic_path-2}
      \end{tabular}
    \end{center}
    \caption{A simplification of \cref{pick_up_two}}
    \label{generic_path}
  \end{figure}
  \begin{compactitem}
    \item If $\mu_w$ also intersects column $i$, then $P_a$ can immediately return to column $i$ if necessary (if $\mu_v$ was a horizontal edge) and collect $\mu_w$. The only care that needs to be taken in this case occurs when $\mu_v$ and $\mu_w$ are a pair of horizontal and vertical edges that share an endpoint.  When this happens, the initial traversal on the boundary should be done (clockwise or counterclockwise) so that the vertical edge appears first in $P_a$.

>>>>

\title{\MakeUppercase{Linear Versus Centred Colouring via Pseudogrids}\thanks{This research was partly funded by NSERC.}}
\author{Prosenjit~Bose\thanks{School of Computer Science, Carleton University},\quad Vida~Dujmović\thanks{Department of Computer Science and EE, University of Ottawa},\quad Hussein~Houdrouge\footnotemark[2],\quad Mehrnoosh~Javarsineh\footnotemark[2],\quad and Pat~Morin\thanks{School of Computer Science, Carleton University, \email{morin@scs.carleton.ca}}}

@@

\keywords{treedepth; graph colouring; centered colouring; linear colouring}

@@

Any further improvement to \cref{kun_obrien_general2} will either require an improved Excluded Grid Theorem or an entirely new approach. However, no improvement to the Excluded Grid Theorem will be sufficient to establish a linear relationship between the centred and the linear chromatic number.  Indeed, the best possible Excluded Grid Theorem would state that any graph of treewidth $k^2\log k$ contains an $\Omega(k)\times\Omega(k)$ grid minor \cite{robertson.seymour.ea:quickly}, and the preceding argument would only show that $\chicen(G)\in O((\chilin(G))^3\log(\chilin(G)))$.  Even the Excluded Grid Theorem for Planar Graphs states that any planar graph of treewidth $k$ contains an $\Omega(k)\times\Omega(k)$ grid minor
% \cite[(6.2)]{robertson.seymour.ea:quickly}.
\cite{robertson.seymour.ea:quickly}.
Combining this with the argument above and \cref{pseudogrid_lower_bound} shows only that, for any planar graph $G$, $\chicen(G)\in\Theta((\chilin(G))^2)$.

@@

\begin{figure}
  \begin{center}
    \includegraphics[page=2]{figs/pseudogrid}
  \end{center}
  \caption{A $6\times 4$ pseudogrid.  Black vertices are contained in $P_v$ for some vertex $v$ of $G_{6\times 4}$ and white vertices are contained in $P_{vw}$ for some edge of $G_{6\times 4}$. Colour are used to distinguish between cases (Q1) (blue), (Q2) (pink), and (Q3) (gold).}
  \label{pseudogrid_fig}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{ccc}
      \includegraphics[page=2]{figs/outline} &
      \includegraphics[page=3]{figs/outline} &
      \includegraphics[page=4]{figs/outline} \\
      $S_1$ & $S_1\cup S_2$ & $P$
    \end{tabular}
  \end{center}
  \caption{Constructing the sets $S_1$ and $S_2$ and the path $P$.}
  \label{outline}
\end{figure}
\begin{compactenum}

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{ccc}
      \includegraphics[page=1]{figs/bad-example} &
      \includegraphics[page=2]{figs/bad-example} &
      \includegraphics[page=3]{figs/bad-example}
    \end{tabular}
  \end{center}
  \caption{Some examples of $3$- and $4$-element vertex subsets that cannot all be contained in a single path (that must also contain other vertices not pictured).}
  \label{bad_examples}
\end{figure}

@@

  \begin{figure}
    \begin{center}
      \begin{tabular}{ccc}
        \includegraphics[page=1,width=.3\textwidth]{figs/make_disjoint} &
        \includegraphics[page=2,width=.3\textwidth]{figs/make_disjoint} &
        \includegraphics[page=3,width=.3\textwidth]{figs/make_disjoint} \\
        $X_1$ & $X_2$ & $X_3$
      \end{tabular}
    \end{center}
    \caption{The proof of \cref{make_disjoint}. Disks are vertices in $S$ and crosses are the resulting elements of $X$. The large yellow boxes are $B_{r-1}(v)$ for some $v\in S$ and the small mauve and green boxes are $B_{p+1}(x)$ and $B_{p}(x)$, respectively, for some $x\in X$.}
    \label{make_disjoint_fig}
  \end{figure}
  Let $S_2$ be the subset of $S\setminus S_1$ containing those vertices that take part in a $2p$-pair.  For each $2p$-pair $v,w\in S_2$, there exists  $x\in V(\interior_r(G))$ such that $v,w\in \tilde{B}_{p}(x)$. We include $x$ in $X_2$ and say that $x$ \defin{covers} $v$ and $w$.

@@

\begin{proof}
  Fix a real number $c\ge 1$ such that $d=cr^4\log r$ is an integer.
  We begin by applying \cref{one_colour_per_object} to obtain the colouring $\phi:\VE(G_k)\to\varphi(V(G))\cup\{\perp\}$ in which $\phi^{-1}(\alpha)\ge d$ for each $\alpha\in\varphi(V(G))$.  Observe that it is now sufficient to find a $2|\varphi(V(G))|$-element subset $Q\subseteq\VE(\interior_r(G_k))$ such that
  \begin{compactenum}[(i)]
    \item $|\phi^{-1}(\alpha)\cap Q|= 2$ for each $\alpha\in\varphi(V(G))$ and
    \item $|\VE(G_r(\mu))\cap Q|\le 2$ for each $\mu\in Q$.
  \end{compactenum}
  Indeed, with such a $Q$ we can obtain $S$ by taking one vertex of colour $\phi(\mu)$ from $P_\mu$ for each $\mu\in Q$.

@@

  For each $\alpha\in X$, let $\Phi_\alpha:=\phi^{-1}(\alpha)\cap \left(\bigcup_{\mu\in Y}\VE(G_{2r+1}(\mu))\right)$.  In words, $\Phi_\alpha$ consists of all objects of colour $\alpha$ that were eliminated from consideration because they are too close to an element of $Y$. We claim that, for each $\alpha\in X$, $|\Phi_\alpha|\ge d-\vol{2r+1}$.  Otherwise $|\phi^{-1}(\alpha)\setminus\Phi_\alpha| > \vol{2r+1}$. Then, when the colour $\alpha$ was considered during the first round, we could have taken $\mu_1$ to be any object in $\phi^{-1}(\alpha)\setminus\Phi_\alpha$ and taken $\mu_2$ to be any object in $\phi^{-1}(\alpha)\setminus\Phi_\alpha\setminus B_{2r+1}(\mu_1)$. (Recall that $\vol{2r+1}\ge |B_{2r+1}(\mu_1)|$, by definition.)  Thus, the first round would have been an $\alpha$-success, contradicting the fact that $\alpha\in X$.

@@

  Comparing \cref{weight_product,probability} we see that both quantities decrease exponentially in $|\Phi_\alpha|$ but \cref{weight_product} does so at a rate that can be controlled with $\tau$.
  Taking $\tau = {32\vol{7r}^2}$, and using \cref{weight_product,probability}, we find that
  \[
     \Pr\left(E_\alpha\right) \le w(E_\alpha)\prod_{\beta\in\Gamma_\alpha}(1-w(E_\beta))
  \]
  provided that
  \begin{equation}
    |\Phi_\alpha| \ge {32\vol{7r}}\big(\ln({32\vol{7r}}+1) + {2\vol{2r+1}}\ln(2|\Phi_\alpha|)\big) \enspace . \label{critical}
  \end{equation}
  The right hand side of this last equation is at most $c'(r^2\log r+r^4\log|\Phi_\alpha|) \le c''r^4\log|\Phi_\alpha|$ for some absolute constants $c'$ and $c''$.  Thus, it suffices to choose $c$ so that
  \[
    \frac{|\Phi_\alpha|}{\log|\Phi_\alpha|}
    \ge \frac{(d-\vol{r})}{\log(d-\vol{r})}
    \ge \frac{(d-\vol{r})}{\log d}
    = \frac{cr^4\log r-\vol{r}}{\log(cr^4\log r)}
    \ge \frac{cr^4\log r-\vol{r}}{\log(cr^5)}
    \ge \frac{cr^4\log r}{2\log(cr^5)}
    \ge c''r^4 \enspace .
  \]
  Rewriting the last inequality gives $c-2\log_r c\ge 10c''$.  Since $r\ge 10$, $c-2\log_r c\ge c/2$, for all $c\ge 1$.  So \eqref{critical} holds for all $c\ge 20c''$.

@@

We are also grateful to Omer Angel for pointing out that the problem on a $k\times k$ grid (as opposed to pseudogrid) has a simple solution, illustrated in \cref{omer}: Partition the grid into $2\times 2$ grids and repeatedly delete a subgrid if it contains a vertex whose colour is unique.  If this causes the grid to become disconnected, keep only the largest remaining component and continue.  This process deletes at most $4\epsilon k$ vertices.  For $\epsilon < 1/4$, this implies that the graph that remains at the end of this process has at least $k^2/2$ vertices.  In particular, the graph that remains is non-empty and every colour that appears in this graph is used on at least two vertices.  Finally, this graph has a Hamiltonian path (which can be constructed by traversing the boundary of a certain tree contained in the dual graph).  Unfortunately, this solution breaks down immediately for pseudogrids since a pseudogrid is not necessarily Hamiltonian, even before removing some of its vertices and edges.

@@

  Although this is a simplification of the problem, we first describe how to solve it.  For this simpler problem, we may assume that each of $\mu_s$, $\mu_v$, $\mu_w$, and $\mu_t$ is an edge of $G_a$ since we can replace any vertex of $G_a$ with one of its incident edges. Refer to \cref{generic_path}. Beginning at $\mu_s$, $P_a$ can traverse the boundary of $G_a$ until reaching the first column $i$ that contains an endpoint of $\mu_v$ or $\mu_w$, then vertically in this column to collect $\mu_v$ (say). What happens next depends on whether or not $\mu_w$ also intersects column $i$.
  \begin{figure}[htbp]
    \begin{center}
      \begin{tabular}{cc}
        \includegraphics[page=1]{figs/generic_path} &
        \includegraphics[page=2]{figs/generic_path}
      \end{tabular}
    \end{center}
    \caption{A simplification of \cref{pick_up_two}}
    \label{generic_path}
  \end{figure}
  \begin{compactitem}
    \item If $\mu_w$ also intersects column $i$, then $P_a$ can immediately return to column $i$ if necessary (if $\mu_v$ was a horizontal edge) and collect $\mu_w$. The only care that needs to be taken in this case occurs when $\mu_v$ and $\mu_w$ are a pair of horizontal and vertical edges that share an endpoint.  When this happens, the initial traversal on the boundary should be done (clockwise or counterclockwise) so that the vertical edge appears first in $P_a$.
